Enzo: An Adaptive Mesh Refinement Code for Astrophysics

### 2013-05-19 code-tests.tex
Reynolds edits.

This cloud is subject to radiation from a point source at the center
of the $y=0$ boundary with an ionizing photon luminosity
$\dot{N}_\gamma = 3 \times 10^{51}$ photons s$^{-1}$, corresponding to
a flux $F_0 = 10^6 \unit{photons s}^{-1} \unit{cm}^{-2}$ at the clump
surface closest to the radiation source.  The radiation source has a
spectrum of a $T = 10^5 \unit{K}$ blackbody, and we use four energy
groups with the following mean energies and relative luminosities:
$E_i = (17.98, 31.15, 49.09, 76.98) \unit{eV}, L_i/L = (0.23, 0.36,
0.24, 0.06)$ that are optimized to reduce errors in the solutions with
a full spectrum and energy discretization \citep{Mirocha12}.
\citep[Note that this choice of energy groups is different
from][]{Wise11_Moray}.  We use a minimum angular resolution of 10 rays
per cell and a constant radiative transfer timestep of 25 kyr.  Figure
\ref{fig:shadowing} depicts the clump after at $t = 15$ Myr with the
outer layers expanding after being photo-heated.  It also shows the
sharp shadowing effects of the dense clump in the neutral fraction
plot that is representative of ray tracing techniques.

>>>>

This cloud is subject to radiation from a point source at the center
of the $x=0$ boundary with an ionizing photon luminosity
$\dot{N}_\gamma = 3 \times 10^{51}$ photons s$^{-1}$, corresponding to
a flux $F_0 = 10^6 \unit{photons s}^{-1} \unit{cm}^{-2}$ at the clump
surface closest to the radiation source.  The radiation source has a
spectrum of a $T = 10^5 \unit{K}$ blackbody, and we use four energy
groups with the following mean energies and relative luminosities:
$E_i = (17.98, 31.15, 49.09, 76.98) \unit{eV}, L_i/L = (0.23, 0.36,
0.24, 0.06)$ that are optimized to reduce errors in the solutions with
a full spectrum and energy discretization \citep{Mirocha12}.
\citep[Note that this choice of energy groups is different
from][]{Wise11_Moray}.  We use a minimum angular resolution of 10 rays
per cell and a constant radiative transfer timestep of 25 kyr.  Figure
\ref{fig:shadowing} depicts the clump after at $t = 15$ Myr with the
outer layers expanding after being photo-heated.  It also shows the
sharp shadowing effects of the dense clump in the neutral fraction
plot that is representative of ray tracing techniques.

### 2013-05-19 numerical-amr.tex
Reynolds edits.

Each grid patch in \enzo\ contains arrays of values for baryon and particle quantities.   Grids are partitioned into a core of \emph{active zones} and a surrounding layer of \emph{ghost zones}, as shown in Figure~\ref{fig:amr_hierarchy}.  The active zones store field values and ghost zones are used to temporarily store values that have been obtained directly from neighboring grids or interpolated from a parent grid.  These zones are necessary to accommodate the computational stencil of the (magneto)hydrodynamics solvers (Sections~\ref{sec.hydro.ppm} and~\ref{sec.hydro.zeus}) and the gravity solver (Section~\ref{sec.gravity}).  The PPM and Zeus hydro solvers require 3 layers of ghost zones,  and the gravity solver requires 6 (although only for the density and potential fields).  This can lead to significant memory and computational overhead, particularly for smaller grid patches at high levels of refinement.

@@

Timesteps are determined on each level as described in \ref{sec.timestepping}, and the hierarchy is advanced on
a level-by-level basis in a W-Cycle.    Beginning with the coarsest level, $l$, all grids on that level are advanced one
timestep.  Then one timestep is taken on all grids at the next level of refinement, $l+1$, and so on until the finest level is resolved. The finest level is then advanced, using as many steps as it takes to reach the level immediately above.  The finest level is then synched to the level above it, which then proceeds foreward in time one more step.  This is shown graphically in Figure~\ref{fig:wcycle}

@@

The basic control algorithm looks very much like any used in a grid-based code of a single resolution, and can be written schematically as:
\begin{verbatim}
InitializeHierarchy
While (Time < StopTime)
begin
  dt = ComputeTimeStep(0)
  EvolveLevel(0, dt)
  Time = Time + dt
  CheckForOutput(Time)
end
\end{verbatim}
The AMR control algorithm is contained within the recursive
EvolveLevel algorithm, which takes the level on which to operate as an
argument.  This looks something like the following pseudo-code:
\begin{verbatim}
EvolveLevel(level)
begin
  SetBoundaryValues
  while (Time < ParentTime)
  begin
    dt = ComputeTimeStep(level)
    SolveHydroEquations(dt)
    Time = Time + dt
    SetBoundaryValues
    EvolveLevel(level+1, dt)
    FluxCorrection
    Projection
    RebuildHierarchy(level+1)
  end
end
\end{verbatim}

@@

In one dimension, such functions are well known; examples include
donor cell, Van Leer and piecewise parabolic interpolation, which are
first, second and third-order accurate respectively
\citep[e.g.,][]{Stone92a}.  However, we require fully
multi-dimensional interpolation functions, and so focus on
second-order schemes since they are tractable and relatively accurate.
\footnote{We could, in principle, build a multi-dimensional
  interpolator out of a one-dimensional interpolation function;
  however, at higher order, this would not necessarily satisfy our
  constraint regarding not introducing new extrema in interpolated quantities.}  In Appendix~\ref{app:interpolation}, we
explicitely write down functions in one, two and three dimensions, but
here we discuss the general design considerations.

In two dimensions, it is relatively easy to build a linear, monotonic,
conservative function by requiring, first, that the function at the
center of the cell return the cell-centered quantitiy (and since it is
linear, this implies that an average over the cell will also be equal
to this value), and second, by treating the cell-diagonals as
one-dimensional Van Leer-type functions.  The reason to focus on the
cell diagonals is that for a linear function, the extrema will occur at the cell 
corners, which are in turn linked by the diagonals.  This process can be done in
two dimensions since there are three constraints (one at the center and one along
each diagonal), and three parameters in a two-dimensional linear
function.

@@

Given the grid and its boundary conditions, the process of solving
equations~(\ref{eq:mass})--(\ref{eq:total_energy}) can be detached
from the rest of the SAMR technique, so that it is possible to
experiment with a number of different methods.  The restrictions are
that the method should be flux-conservative, or at least produce an
accurate estimate of the flux of conserved quantities at cell
interfaces.  This is required in the flux-correction step (see
\ref{sec:flux_correction}).  There is also the issue of centering,
which refers to the position of physical quantities within a cell (the
primary types are cell-centered, face-centered, and node-centered). We
will employ the direct Eulerian version of the piecewise
parabolic method \citep{1984JCoPh..54..174C,1995CoPhC..89..149B}, which
assumes cell-centering for all quantities (actually they are
cell-averages).  However, it is also possible to use other schemes,
which employ different centering for different variables
\citep[e.g.,][]{Stone92a}.  The interpolation step, described in
Section~\ref{sec:interpolation}, must correctly interpolate with the
selected centering.

@@

Since the addition of more highly refined grids is adaptive, the conditions for refinement must be specified. Many different criteria can be simultaneously specified for a given simulation, and these refinement criteria are discussed in more detail in the following section (Section~\ref{sec:refinement_criteria}).   Once all cells in a given grid that require additional refinement have been flagged, rectangular boundaries are determined that minimally encompass the flagged cells.  This is done using an iterative machine-vesion technique as described in \red{(REF)}.

We first find the smallest region (or ``proto-subgrid") that covers all of the refined cells on a grid.  We compute an efficiency $\epsilon$ for this grid, defined as the ratio of flagged cells to total cells.  If this efficiency is sufficiently larger (typically larger than 30\%), or if the grid is small enough (typically less than $4^d$ coarse cells, where d is the dimensionality of the simulation) then we adopt the region as a new grid and move on to the next region (if any).  If the region is not acceptable, then we split it into two using the following algorithm.  First, we determine the longest dimension and compute the signature along that dimension.  The signature is found by summing the number of refined cells with the same index along the longest dimension.  Edges are typically indicates by inflection points in this signature, so we calculate the second derivative using a simple three-point finite difference estimator and look for the maximum in the second derivative, with ties going to the index closest to the center.  If no inflection points are found along the longest exis, the next-longest axis is searched.  If no inflection points are found, the longest axis is cut in half.  The original region is then split into two new regions, or proto-subgrids.  This process is then re-applied to each of the new regions.

This algorithm is not necessarily optimal, but experience shows it does a reasonable job of generating an efficient covering of the refined cells.  Note that some cells are unnecessarily refined with the SAMR approach, resulting in wasted resources (unlike in cell-splitting AMR schemes).  However, it has a number of advantages.  First, traditional grid-based hydro method can be simply added with only minor modifications.  Second, it is easier to accommodate larger stencils (as in PPM, which requires three boundary zones), and third, it can be more efficient for some machines since there are more predictable and linear memory accesses, which more efficiently use a CPU's cache.

>>>>

Each grid patch in \enzo\ contains arrays of values for baryon and particle quantities.   Grids are partitioned into a core of \emph{active zones} and a surrounding layer of \emph{ghost zones}, as shown in Figure~\ref{fig:amr_hierarchy}.  \red{(Move this figure to here?)}  The active zones store field values and ghost zones are used to temporarily store values that have been obtained directly from neighboring grids or interpolated from a parent grid.  These zones are necessary to accommodate the computational stencil of the (magneto)hydrodynamics solvers (Sections~\ref{sec.hydro.ppm} and~\ref{sec.hydro.zeus}) and the gravity solver (Section~\ref{sec.gravity}).  The PPM and Zeus hydro solvers require 3 layers of ghost zones,  and the gravity solver requires 6 (although only for the density and potential fields).  This can lead to significant memory and computational overhead, particularly for smaller grid patches at high levels of refinement.

@@

Timesteps are determined on each level as described in Section~\ref{sec.timestepping}, and the hierarchy is advanced on
a level-by-level basis in a W-Cycle.    Beginning with the coarsest level, $l$, all grids on that level are advanced one
timestep.  Then one timestep is taken on all grids at the next level of refinement, $l+1$, and so on until the finest level is resolved. The finest level is then advanced, using as many steps as it takes to reach the level immediately above.  The finest level is then synched to the level above it, which then proceeds foreward in time one more step.  This is shown graphically in Figure~\ref{fig:wcycle}.

@@

The basic control algorithm looks very much like any used in a grid-based code of a single resolution, and can be written schematically as:
\begin{verbatim}
InitializeHierarchy
While (Time < StopTime)
begin
  dt = ComputeTimeStep(0)
  EvolveLevel(0, dt)
  Time = Time + dt
  CheckForOutput(Time)
end
\end{verbatim}
The AMR control algorithm is contained within the recursive
EvolveLevel algorithm, which takes the level on which to operate as an
argument.  This looks something like the following pseudo-code:
\begin{verbatim}
EvolveLevel(level)
begin
  SetBoundaryValues
  while (Time < ParentTime)
  begin
    dt = ComputeTimeStep(level)
    SolveHydroEquations(dt)
    SolveOtherEquations(dt)
    Time = Time + dt
    SetBoundaryValues
    EvolveLevel(level+1, dt)
    FluxCorrection
    Projection
    RebuildHierarchy(level+1)
  end
end
\end{verbatim}

@@

In one dimension, such functions are well known; examples include
donor cell, Van Leer and piecewise parabolic interpolation, which are
first, second and third-order accurate respectively
\citep[e.g.,][]{Stone92a}.  However, we require fully
multi-dimensional interpolation functions, and so focus on
second-order schemes since they are tractable and relatively accurate.
\footnote{We could, in principle, build a multi-dimensional
  interpolator out of a one-dimensional interpolation function;
  however, at higher order, this would not necessarily satisfy our
  constraint regarding not introducing new extrema in interpolated quantities.}  In Appendix~\ref{app:interpolation}, we
explicitly write down functions in one, two and three dimensions, but
here we discuss the general design considerations.

In two dimensions, it is relatively easy to build a linear, monotonic,
conservative function by requiring, first, that the function at the
center of the cell return the cell-centered quantity (and since it is
linear, this implies that an average over the cell will also be equal
to this value), and second, by treating the cell-diagonals as
one-dimensional Van Leer-type functions.  The reason to focus on the
cell diagonals is that for a linear function, the extrema will occur at the cell 
corners, which are in turn linked by the diagonals.  This process can be done in
two dimensions since there are three constraints (one at the center and one along
each diagonal), and three parameters in a two-dimensional linear
function.

@@

Given the grid and its boundary conditions, the process of solving
equations~(\ref{eq:mass})--(\ref{eq:total_energy}) can be detached
from the rest of the SAMR technique, so that it is possible to
experiment with a number of different methods.  The restrictions are
that the method should be flux-conservative, or at least produce an
accurate estimate of the flux of conserved quantities at cell
interfaces.  This is required in the flux-correction step (see
\ref{sec:flux_correction}).  There is also the issue of centering,
which refers to the position of physical quantities within a cell (the
primary types are cell-centered, face-centered, and node-centered). In this description we 
will employ the direct Eulerian version of the piecewise
parabolic method \citep{1984JCoPh..54..174C,1995CoPhC..89..149B}, which
assumes cell-centering for all quantities (actually they are
cell-averages).  However, it is also possible to use other schemes,
which employ different centering for different variables
\citep[e.g.,][]{Stone92a}.  The interpolation step, described in
Section~\ref{sec:interpolation}, must correctly interpolate with the
selected centering.

@@

Since the addition of more highly refined grids is adaptive, the conditions for refinement must be specified. Many different criteria can be simultaneously specified for a given simulation, and these refinement criteria are discussed in more detail in the following section (Section~\ref{sec:refinement_criteria}).   Once all cells in a given grid that require additional refinement have been flagged, rectangular boundaries are determined that minimally encompass the flagged cells.  This is done using an iterative machine-vision technique, as described in \red{(REF)}.

We first find the smallest region (or ``proto-subgrid") that covers all of the refined cells on a grid.  We compute an efficiency $\epsilon$ for this grid, defined as the ratio of flagged cells to total cells.  If this efficiency is sufficiently larger (typically larger than 30\%), or if the grid is small enough (typically less than $4^d$ coarse cells, where d is the dimensionality of the simulation) then we adopt the region as a new grid and move on to the next region (if any).  If the region is not acceptable, then we split it into two using the following algorithm.  First, we determine the longest dimension and compute the signature along that dimension.  The signature is found by summing the number of refined cells with the same index along the longest dimension.  Edges are typically indicated by inflection points in this signature, so we calculate the second derivative using a simple three-point finite difference estimator and look for the maximum in the second derivative, with ties going to the index closest to the center.  If no inflection points are found along the longest exis, the next-longest axis is searched.  If no inflection points are found, the longest axis is cut in half.  The original region is then split into two new regions, or proto-subgrids.  This process is then re-applied to each of the new regions.

This algorithm is not necessarily optimal, but experience shows it does a reasonable job of generating an efficient covering of the refined cells.  Note that some cells are unnecessarily refined with the SAMR approach, resulting in wasted resources (unlike in cell-splitting AMR schemes).  However, it has a number of advantages.  First, traditional grid-based hydro methods can be simply added with only minor modifications.  Second, it is easier to accommodate larger stencils (as in PPM, which requires three boundary zones), and third, it can be more efficient for some machines since there are more predictable and linear memory accesses, which more efficiently use a CPU's cache.

### 2013-05-19 numerical-chemistry.tex
Reynolds edits.

The chemical reaction equation network is technically challenging to
solve due to the huge range of reaction timescales involved -- the
characteristic creation and destruction timescales of the various
species and reactions can differ by many orders of magnitude.  As a
result, the set of rate equations is extremely stiff, and an explicit
scheme for integration of the rate equations can be exceptionally
costly if small enough timesteps are taken to keep the network stable.
This makes an implicit scheme strongly preferable for such a set of
equations.  However, an implicit scheme typically require an iterative
procedure to converge, and for large networks (such as this one) an
implicit method can be very time-consuming, making it undesirable for
a large, three-dimensional simulation.

\enzo\ solves the rate equations using a method based on a backwards
differencing formula (BDF) in order to provide a stable and accurate
solution. This technique is optimized by taking the chemical
intermediaries \Hm and \HHp, which have large rate coefficients and
low concentrations, and grouping them into a separate category of
equations.  Due to their fast reactions, these species are very
sensitive to small changes in the more abundant species.  However, due
to their low overall concentrations, they do not significantly
influence the abundance of species with higher concentrations.
Therefore, reactions involving these two species can be decoupled from
the rest of the network and treated independently.  In fact, \Hm~and
\HHp~are treated as being in equilibrium at all times, independent of
the other species and the hydrodynamic variables.  This allows a large
speedup in solution as the BDF scheme is then applied only to the
slower 7- or 10-species network (depending on whether deuterium is
included or not) on timescales closer to those required by the
hydrodynamics of the simulation.  Even so, the accuracy and stability
of the scheme is maintained by subcycling the rate solver over a
single hydrodynamic timestep.  These subcycle timesteps are determined
so that the estimated fractional change in the electron concentration
is limited to no more than $10\%$ per timestep.  Given this
subcycling, the method is not truly implicit over an entire
hydrodynamical timestep, and thus not truly a BDF method -- however,
it uses the formalism and is implicit over subcycle timesteps.

It is important to note the regime in which this model is valid.
According to \citet{abel97} and \citet{anninos97},
the reaction network is valid for temperatures between $10^0 - 10^8$
K.  The original model discussed in these two references is only valid
up to $n_H \sim 10^4$~cm$^{-3}$.  However, addition of the 3-body
\HH~formation process (equation 20 in Table~\ref{table.collisional})
allows correct modeling of the chemistry of the gas up until the point
where collisionally-induced emission from molecular hydrogen becomes
an important cooling process, which occurs at $n_{\rm H} \sim
10^{14}$~cm$^{-3}$.  A further concern is that the optically thin
approximation for radiative cooling eventually breaks down, which occurs before
$n_{\rm H} \sim 10^{16} - 10^{17}$~cm$^{-3}$ in gas of primordial composition.  Beyond this point,
modifications the cooling function that take into account the
non-negligible opacity in the gas must be made, as discussed by
\citet{2004MNRAS.348.1019R}, and was put into \enzo\ for the work
published in \citep{2009Sci...325..601T,2009PhDT.........5T}.  Even
with these modifications, a completely correct description of the
cooling of primordial gas requires some form of radiation transport,
which will greatly increase the cost of the simulations.

>>>>

The chemical reaction equation network is technically challenging to
solve due to the huge range of reaction timescales involved -- the
characteristic creation and destruction timescales of the various
species and reactions can differ by many orders of magnitude.  As a
result, the set of rate equations is extremely stiff, and an explicit
scheme for integration of the rate equations can be exceptionally
costly if small enough timesteps are taken to keep the network stable.
This makes an implicit scheme strongly preferable for such a set of
equations.  However, an implicit scheme typically requires an iterative
procedure to converge, and for large networks (such as this one) an
implicit method can be very time-consuming, making it undesirable for
a large, three-dimensional simulation.

\enzo\ solves the rate equations using a method based on a first-order-accurate, implicit
Euler formulation in order to provide a stable and accurate
solution. This technique is optimized by taking the chemical
intermediaries \Hm and \HHp, which have large rate coefficients and
low concentrations, and grouping them into a separate category of
equations.  Due to their fast reactions, these species are very
sensitive to small changes in the more abundant species.  However, due
to their low overall concentrations, they do not significantly
influence the abundance of species with higher concentrations.
Therefore, reactions involving these two species can be decoupled from
the rest of the network and treated independently.  In fact, \Hm~and
\HHp~are treated as being in equilibrium at all times, independent of
the other species and the hydrodynamic variables.  This allows a large
speedup in solution as the implicit Euler scheme is then applied only to the
slower 7- or 10-species network (depending on whether deuterium is
included or not) on timescales closer to those required by the
hydrodynamics of the simulation.  Even so, the accuracy and stability
of the scheme is maintained by subcycling the rate solver over a
single hydrodynamic timestep.  These subcycle timesteps are determined
so that the estimated fractional change in the electron concentration
is limited to no more than $10\%$ per timestep.  Given this
subcycling, the method is not truly implicit over an entire
hydrodynamical timestep  -- however,
it uses the formalism and is implicit over subcycle timesteps.

It is important to note the regime in which this model is valid.
According to \citet{abel97} and \citet{anninos97},
the reaction network is valid for temperatures between $10^0 - 10^8$
K.  The original model discussed in these two references is only valid
up to $n_H \sim 10^4$~cm$^{-3}$.  However, addition of the 3-body
\HH~formation process (equation 20 in Table~\ref{table.collisional})
allows correct modeling of the chemistry of the gas up until the point
where collisionally-induced emission from molecular hydrogen becomes
an important cooling process, which occurs at $n_{\rm H} \sim
10^{14}$~cm$^{-3}$.  A further concern is that the optically thin
approximation for radiative cooling eventually breaks down, which occurs before
$n_{\rm H} \sim 10^{16} - 10^{17}$~cm$^{-3}$ in gas of primordial composition.  Beyond this point,
modifications to the cooling function that take into account the
non-negligible opacity in the gas must be made, as discussed by
\citet{2004MNRAS.348.1019R}, and was put into \enzo\ for the work
published in \citep{2009Sci...325..601T,2009PhDT.........5T}.  Even
with these modifications, a completely correct description of the
cooling of primordial gas requires some form of radiation transport,
which will greatly increase the cost of the simulations.

### 2013-05-19 numerical-mhd-ct.tex
Reynolds edits.

The Constrained Transport (CT) method \citep{Evans88, Balsara99} for
magnetohydrodynamic (MHD) evolution employs this same vector property
to evolve the magnetic field in a manner that preserves $\Divb=0$ to
to limits of machine precision.  The electric field is computed using
the fluxes from the Riemann solver.  The curl of that electric field
is then used to update the magnetic field.  The advantage of this
method is that it preserves \Divb\ to machine precision.  The drawback
is twofold: increased algorithm complexity, as will be described in
this section; and since only the update of the magnetic field is
divergence free, any monopoles created by other numerical sources
persist.

@@

Divergence-free AMR is somewhat more complex than the AMR employed
elsewhere in Enzo.  First, the interpolation must be constrained to be
divergence free.  Thus, all three face-centered field components are
interpolated in concert.  Second, any magnetic field information from
previous time steps must be included in the interpolation, making the
interpolation more complex than the simple parent-child relation used
for other fields.  Third, the flux correction involves more possible
grid relations than traditional AMR.  In order to circumvent this last
complexity, the electric field is projected from fine grids to parent
grids (rather than the magnetic field), and is then used to re-update
the parent magnetic field.  This is described in detail in
\citet{Balsara99} and \citet{Collins10}.

>>>>

The Constrained Transport (CT) method \citep{Evans88, Balsara99} for
magnetohydrodynamic (MHD) evolution employs this same vector property
to evolve the magnetic field in a manner that preserves $\Divb=0$ to
limits of machine precision.  The electric field is computed using
the fluxes from the Riemann solver.  The curl of that electric field
is then used to update the magnetic field.  The advantage of this
method is that it preserves \Divb\ to machine precision.  The drawback
is twofold: increased algorithm complexity, as will be described in
this section; and since only the update of the magnetic field is
divergence free, any monopoles created by other numerical sources
persist.

@@

Divergence-free AMR is somewhat more complex than the AMR employed
elsewhere in Enzo.  First, the interpolation must be constrained to be
divergence free.  Thus, all three face-centered field components are
interpolated in concert.  Second, any magnetic field information from
the previous time step must be included in the interpolation, making the
interpolation more complex than the simple parent-child relation used
for other fields.  Third, the flux correction involves more possible
grid relations than traditional AMR.  In order to circumvent this last
complexity, the electric field is projected from fine grids to parent
grids (rather than the magnetic field), and is then used to re-update
the parent magnetic field.  This is described in detail in
\citet{Balsara99} and \citet{Collins10}.

### 2013-05-19 numerical-overview.tex
Many minor changes.  No more undefined references!  Added appendix file.
Reynolds edits.

%---------------------------------------------------------------------------%
\subsection{Chemistry, radiation and cooling}\label{sec.ov.chem}

@@

\enzo\ solves the rate equations using a method based on a backwards differencing 
formula (BDF) in order to provide a stable and accurate solution.  This technique is
optimized by taking the chemical intermediaries $H^-$ and $H_2^+$, which have 
large rate coefficients and low concentrations, and grouping them into a separate
category of equations.  Due to their fast reactions these species are very sensitive
to small changes in the more abundant species.  However, due to their low overall
concentrations, they do not significantly influence the abundance of species with
higher concentrations.  Therefore, reactions involving these two species can be
decoupled from the rest of the network and treated independently.  In fact, $H^-$ 
and $H_2^+$ are treated as being in equilibrium at all times, independent of 
the other species and the hydrodynamic variables.  This allows a large speedup
in solution as the BDF scheme is then applied only to the slower 7-species network
on timescales closer to those required by the hydrodynamics of the simulation.
Even so, the accuracy and stability of the scheme is maintained by subcycling the 
rate solver over a single hydrodynamic timestep.  These subcycle timesteps are 
determined so that the maximum fractional change in the electron concentration is
limited to no more than $10\%$ per timestep.

>>>>

%---------------------------------------------------------------------------%
\subsection{Chemistry, radiation and cooling}
\label{sec.ov.chem}

@@

\enzo\ solves the rate equations using a method based on an implicit
Euler formulation in order to provide a stable and accurate solution.  This technique is
optimized by taking the chemical intermediaries $H^-$ and $H_2^+$, which have 
large rate coefficients and low concentrations, and grouping them into a separate
category of equations.  Due to their fast reactions these species are very sensitive
to small changes in the more abundant species.  However, due to their low overall
concentrations, they do not significantly influence the abundance of species with
higher concentrations.  Therefore, reactions involving these two species can be
decoupled from the rest of the network and treated independently.  In fact, $H^-$ 
and $H_2^+$ are treated as being in equilibrium at all times, independent of 
the other species and the hydrodynamic variables.  This allows a large speedup
in solution as the implicit scheme is then applied only to the slower 7-species network
on timescales closer to those required by the hydrodynamics of the simulation.
Even so, the accuracy and stability of the scheme is maintained by subcycling the 
rate solver over a single hydrodynamic timestep.  These subcycle timesteps are 
determined so that the maximum fractional change in the electron concentration is
limited to no more than $10\%$ per timestep.

### 2013-05-19 numerical-particle.tex
Reynolds edits.

\enzo\ uses a particle-mesh N-body method to calculate the dynamics of
collisionless systems \citep{Hockney88}.  This method follows
trajectories of a representative sample of individual particles that
sample the phase space of the dark matter distribution, and is much
more efficient than a direct solution of the Boltzmann equation in
essentially all astrophysical situations for the levels of accuracy
that are required for simulations of structure formation.  As
described earlier, the gravitational potential is computed by solving
the elliptic Poisson's equation (Eq.~\ref{eq:potential}), and
differencing the potential to find accelerations, which are then
interpolated back to particles.  This acceleration is time-centered
(because the underlying gravitating mass field is approximately
time-centered), and so we have accelerations $\myvec{g}^{n+1/2}$ for
each particle.  These are used to update the particle positions and
velocities starting from $\myvec{x}^n$ and $\myvec{v}^n$ using a
standard drift-kick-drift technique:

>>>>

\enzo\ uses a particle-mesh N-body method to calculate the dynamics of
collisionless systems \citep{Hockney88}.  This method follows
trajectories of a representative sample of individual particles that
sample the phase space of the dark matter distribution, and is much
more efficient than a direct solution of the Boltzmann equation in
essentially all astrophysical situations for the levels of accuracy
that are required for simulations of structure formation.  As
described earlier, the gravitational potential is computed by solving
the elliptic Poisson's equation (Eq.~\ref{eq:potential}), and
differencing the potential to find accelerations, which are then
interpolated back to particles.  This acceleration is time-centered
(because the underlying gravitating mass field is approximately
time-centered), and so we have accelerations $\myvec{g}^{n+1/2}$ for
each particle.  These are used to update the particle positions and
velocities starting from $\myvec{x}^n$ and $\myvec{v}^n$ using a
standard drift-kick-drift technique \citep{Hockney88}:

### 2013-05-19 numerical-ppm.tex
Reynolds edits.

We now restrict ourselves to the solution of Equations
(\ref{eq:mass1d})--(\ref{eq:total_energy1d}), plus transport, in one
dimension.  In the Direct Eulerian version of PPM, this is
accomplished by a three-step process.  First, we compute effective
left and right states at each grid boundary by constructing a
piecewise parabolic description of the primary variables ($\rho$, $u$
and $E$) and then averaging over the regions corresponding to the
distance each characteristic wave can travel ($u$, $u-c$ and $u+c$).
Second, a Riemann problem is solved using these effective left and
right states, and finally the fluxes are computed based on the
solution to this Riemann problem.  This is described in detail in
\citet{1984JCoPh..54..174C}, but we will briefly outline the procedure
here both for completeness and to put the changes we will make in
context.

@@

We first construct monotonic piecewise 
parabolic (third order) interpolations in one dimension
for each of $p$, $\rho$, and $v$  The pressure is determined from
Equation~(\ref{eq:eq_of_state}), the equation of state.
%We have dropped the subscript b in this and the following two sections because we will be referring solely to baryonic quantities.  
The interpolation formula for some quantity $q$ is given by:
%
\begin{eqnarray}
q_j(x) & = &  q_{L,j} + \tilde{x}(\Delta q_j + q_{6,j}(1-\tilde{x})), \\
\tilde{x}      & \equiv & {x - x_{j-1/2} \over \Delta x_j}, \qquad
             x_{j-1/2} \leq x \leq x_{j+1/2}. \nonumber
\end{eqnarray}
%
This is Equation~(1.4) of CW84 (in the radial rather than mass coordinate as is appropriate for the direct rulerian approach). 
%The zones edges in the mass coordinates are $m_{j-1/2}$ and $m_{j+1/2}$ for the left and right edges, respectively.  
The quantities $q_{L,j}$, $\Delta q_j$,
and $q_{6,j}$ can be viewed simply as interpolation constants; however,
they also have more intuitive meanings.  For example, $q_{L,j}$ is the
value of $q$ at the left edge of zone j, while $\Delta q_j$ and $q_{6,j}$ are analogous to the slope and first order correction to the slope of $q$ (see \citet{1984JCoPh..54..174C} for a complete discussion):
\begin{equation}
\Delta q_j \equiv q_{R,j} - q_{L,j} \qquad 
q_{6,j}    \equiv 6\left[q_j - 1/2\left(q_{L,j} + q_{R,j}\right)\right].
\end{equation}

@@

It should be stressed that as long as the parameter ${\eta}_1$ is small enough
the dual energy method {\it will have no dynamical effect}.  We use
${\eta}_1 = 10^{-3}$, which is consistent with the truncation error of the
scheme.  We are now free to select the method by which the internal energy
field variable $e$ is updated so that it will not become contaminated with
errors advected by the total energy formulation but still give the
correct distribution in shocked regions. 
Since we are concerned with the advection of errors, the selection
criterion must look at each cell's local neighbourhood.
In one dimension, this is done with,
\begin{equation}
e = \cases{ (E - \vecv^2/2),& $\rho(E - \vecv^2/2)/
    \max ({\rho}_{j-1}E_{j-1},{\rho}_j E_j,{\rho}_{j+1}E_{j+1}) > {\eta}_2$,\cr
            e,& $\rho(E - \vecv^2/2)/
    \max ({\rho}_{j-1}E_{j-1},{\rho}_j E_j,{\rho}_{j+1}E_{j+1}) < {\eta}_2$.\cr}
    \label{eq:dualselect}
\end{equation}
Thus, ${\eta}_2$ determines when the synchronization (of $e$ with $E$) occurs.
%is a measure of how much error we will allow to be passed
%from neighboring cells.  
Too high a value may mask relatively weak shocks, 
while spurious heating (via contamination) may occur if it is set too low.
After some experimentation,
we have chosen ${\eta}_2 = 0.08$, a somewhat conservative value.
This scheme is optional and is generally only required in large-scale 
cosmological simulations where the gas cools to the expansion of the universe
but large bulk flows develop due to the formation of structure.

>>>>

We now restrict ourselves to the solution of Equations
(\ref{eq:mass1d})--(\ref{eq:total_energy1d}), plus transport, in one
dimension.  In the Direct Eulerian version of PPM, this is
accomplished by a three-step process.  First, we compute effective
left and right states at each grid boundary by constructing a
piecewise parabolic description of the primary variables ($\rho$, $u$
and $E$) and then averaging over the regions corresponding to the
distance each characteristic wave can travel ($u$, $u-c_s$ and
$u+c_s$, where $c_s$ is the sound speed).
Second, a Riemann problem is solved using these effective left and
right states, and finally the fluxes are computed based on the
solution to this Riemann problem.  This is described in detail in
\citet{1984JCoPh..54..174C}, but we will briefly outline the procedure
here both for completeness and to put the changes we will make in
context.

@@

We first construct monotonic piecewise 
parabolic (third order) interpolations in one dimension
for each of $p$, $\rho$, and $v$  The pressure is determined from
Equation~(\ref{eq:eq_of_state}), the equation of state.
%We have dropped the subscript b in this and the following two sections because we will be referring solely to baryonic quantities.  
The interpolation formula for some quantity $q$ is given by:
%
\begin{eqnarray}
q_j(x) & = &  q_{L,j} + \tilde{x}(\Delta q_j + q_{6,j}(1-\tilde{x})), \\
\tilde{x}      & \equiv & {x - x_{j-1/2} \over \Delta x_j}, \qquad
             x_{j-1/2} \leq x \leq x_{j+1/2}. \nonumber
\end{eqnarray}
%
This is Equation~(1.4) of CW84 (in the radial rather than mass coordinate as is appropriate for the direct Eulerian approach). 
%The zones edges in the mass coordinates are $m_{j-1/2}$ and $m_{j+1/2}$ for the left and right edges, respectively.  
The quantities $q_{L,j}$, $\Delta q_j$,
and $q_{6,j}$ can be viewed simply as interpolation constants; however,
they also have more intuitive meanings.  For example, $q_{L,j}$ is the
value of $q$ at the left edge of zone j, while $\Delta q_j$ and $q_{6,j}$ are analogous to the slope and first order correction to the slope of $q$ (see \citet{1984JCoPh..54..174C} for a complete discussion):
\begin{equation}
\Delta q_j \equiv q_{R,j} - q_{L,j} \qquad 
q_{6,j}    \equiv 6\left[q_j - 1/2\left(q_{L,j} + q_{R,j}\right)\right].
\end{equation}

@@

It should be stressed that as long as the parameter ${\eta}_1$ is small enough
the dual energy method {\it will have no dynamical effect}.  We use
${\eta}_1 = 10^{-3}$, which is consistent with the truncation error of the
scheme for grid sizes that are typically used in our simulations.  We are now free to select the method by which the internal energy
field variable $e$ is updated so that it will not become contaminated with
errors advected by the total energy formulation but still give the
correct distribution in shocked regions. 
Since we are concerned with the advection of errors, the selection
criterion must look at each cell's local neighbourhood.
In one dimension, this is done with,
\begin{equation}
e = \cases{ (E - \vecv^2/2),& $\rho(E - \vecv^2/2)/
    \max ({\rho}_{j-1}E_{j-1},{\rho}_j E_j,{\rho}_{j+1}E_{j+1}) > {\eta}_2$,\cr
            e,& $\rho(E - \vecv^2/2)/
    \max ({\rho}_{j-1}E_{j-1},{\rho}_j E_j,{\rho}_{j+1}E_{j+1}) < {\eta}_2$.\cr}
    \label{eq:dualselect}
\end{equation}
Thus, ${\eta}_2$ determines when the synchronization (of $e$ with $E$) occurs.
%is a measure of how much error we will allow to be passed
%from neighboring cells.  
Too high a value may mask relatively weak shocks, 
while spurious heating (via contamination) may occur if it is set too low.
After some experimentation,
we have chosen ${\eta}_2 = 0.08$, a somewhat conservative value.
This scheme is optional and is generally only required in large-scale 
cosmological simulations where the gas cools due to the expansion of the universe
but large bulk flows develop due to the formation of structure.
